# Model choice exercise

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Load the libraries.

```{r load_libraries}
library(ape)
library(phangorn)
```

Load in the sequence data and the neighbour joining tree.

```{r load_msa_tree}
myalignment <- read.dna("ray2000_aligned.fas",format="fasta",as.matrix=TRUE)
mytree  <- nj(dist.dna(myalignment,"TN93"))
```

Now compare different models.

```{r model_test}
myalignment.phydat <- as.phyDat(myalignment) # Convert to a format modeltest understand
myalignment.modeltest <- modelTest(myalignment.phydat,
                             tree=mytree,
                             model = c("JC", "F81", "K80", "HKY", "SYM", "GTR"),
                             G = TRUE,
                             I = TRUE,
                             k = 4,
                             control = pml.control(epsilon = 1e-08, maxit = 3, trace = 1),
                             multicore = FALSE)
myalignment.modeltest
```

Using R, we can look for the model e.g. with the lowest AIC.

```{r}
myalignment.modeltest$Model[myalignment.modeltest$AIC==min(myalignment.modeltest$AIC)]
```

Now obtain a maximum likelihood tree, starting from the neighbour joining tree, and using the model chosen from ```modelTest```.

```{r fit_model}
myalignment.pml <- pml(mytree,myalignment.phydat,model="GTR+I+G",k=4)
# optimise
myalignment.pml <- optim.pml(myalignment.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optInv=TRUE,optGamma=TRUE,optEdge=TRUE)
# display
myalignment.pml
```

This shows the maximum likelihood parameter values. The ML tree is contained in the fit as well.

```{r}
myalignment.pml$tree
```

We can compare our original tree and our new tree numerically using ```treedist```.

```{r}
treedist(myalignment.pml$tree,mytree)
```

This isn't particularly helpful. A comparison of edge lengths is quite striking.

```{r}
sum(myalignment.pml$tree$edge.length)
sum(mytree$edge.length)
```

Now we can perform a bootstrap, starting with our 'best' maximum likelihood tree, and performing nearest neighbour interchanges.

```{r ml_bootstrap,eval=FALSE}
myalignment.pml.bs <- bootstrap.pml(myalignment.pml,bs=100,trees=TRUE,optNni=TRUE)
```

As with the neighbour-joining tree, we can overlay bootstrap supports on our maximum likelihood tree.

```{r plot_ml_bootstrap,eval=FALSE}
plotBS(myalignment.pml$tree,myalignment.pml.bs,type="phylogram",cex=0.5)
```
