# Distances walkthrough

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

First load the libraries you need.

```{r load_library}
library(ape)
library(phangorn)
library(network)
```

Load the multiple sequence alignment using ```read.dna``` from the ```ape``` library.

```{r load_alignment}
myalignment.filename <- "ray2000_pruned_degapped.fas"
myalignment <- read.dna(myalignment.filename,format="fasta",as.matrix=TRUE)
```

Calculate distances using the TN93 model.

```{r tn93}
myalignment.dist.tn93 <- dist.dna(myalignment,model="TN93",as.matrix=TRUE)
```

Plot out the distances.

```{r plot_distances}
hist(myalignment.dist.tn93[lower.tri(myalignment.dist.tn93)],xlab="Distance",ylab="Frequency",main="",col="grey")
```

Calculate a neighbour joining tree.

```{r nj}
myalignment.tn93.nj <- nj(myalignment.dist.tn93)
```

Plot the tree out and add a scale bar.

```{r plot_tree}
plot(myalignment.tn93.nj,"unrooted",cex=0.5)
add.scale.bar(length=0.1)
```

Bootstrap the alignment; we first define a function to make a tree from the alignment, then call ```boot.phylo```.

```{r bootstrap}
maketree <- function(x) nj(dist.dna(x,model="TN93"))
myalignment.tn93.nj.boot <- boot.phylo(myalignment.tn93.nj,myalignment,maketree,B=100,block=3,trees=TRUE,quiet=FALSE,rooted=FALSE)
```

Now we can add the bootstrap supports to the tree.

```{r plot_bootstrap}
plotBS(myalignment.tn93.nj,myalignment.tn93.nj.boot$trees,type="phylogram",cex=0.5)
```

We can also create a network of 'similar' sequences, by defining a threshold distance.

```{r plot_network}
thresh <- 0.01
myalignment.dist.tn93.net <- network(myalignment.dist.tn93<=thresh)
plot(myalignment.dist.tn93.net)
```

