# Getting annotations

Sequences in Genbank also have metadata associated with them. I have included a function, ```read.gbc``` in ```utils.R``` that extracts the accession, country, collection date and host from sequence data.

```{r}
source("utils.R")
```

This function is rather slow at the moment, but works on datasets of less than 500 sequences.

```{r}
annotations <- read.gbc("denv.xml")
```

This function returns a dataframe of the annotations.

```{r}
head(annotations)
```


For those of you interested, here is the code.

```{r,eval=FALSE}
null.to.other <-
function(x,y=NA){
  if(is.null(x)){
    return(y)
  }
  else{
    return(x)
  }
}

read.gbc <- function(fn,verbose=TRUE){
  myxml <- xmlTreeParse(fn)
  xmltop <- xmlRoot(myxml)
  numseq <- length(xmltop)
  accession <- rep("",numseq)
  country <- rep(NA,numseq)
  collectiondate <- rep(NA,numseq)
  host <- rep(NA,numseq)
  for(i in 1:numseq){
    if(verbose){
      print(paste("Processing",i,"of",numseq))
    }
    accession[i] <-  unlist(lapply(getNodeSet(xmltop[[i]],"//INSDSeq/INSDSeq_primary-accession"),xmlValue))[1]
    country[i] <- null.to.other(unlist(lapply(getNodeSet(xmltop[[i]],"//INSDSeq/INSDSeq_feature-table/INSDFeature/INSDFeature_quals/INSDQualifier[INSDQualifier_name='country']/INSDQualifier_value"),xmlValue))[1],NA)
    collectiondate[i] <- null.to.other(unlist(lapply(getNodeSet(xmltop[[i]],"//INSDSeq/INSDSeq_feature-table/INSDFeature/INSDFeature_quals/INSDQualifier[INSDQualifier_name='collection_date']/INSDQualifier_value"),xmlValue))[1],NA)
    host[i] <- null.to.other(unlist(lapply(getNodeSet(xmltop[[i]],"//INSDSeq/INSDSeq_feature-table/INSDFeature/INSDFeature_quals/INSDQualifier[INSDQualifier_name='host']/INSDQualifier_value"),xmlValue))[1],NA)
  }
    return(data.frame(accession,country,collectiondate,host))
}
```