# 'Mugration' exercise

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Load the libraries.

```{r load_libraries} 
library(ape)
library(phangorn)
library(ggtree)
```

Load the FASTA data.

```{r load_data} 
myseqs <- read.dna("H5N1.fas",format="fasta")
```

```{r nj}
mytree  <- nj(dist.dna(myseqs,model="TN93"))
```

Starting with the neighbour joining tree, we reconstruct a maximum likelihood tree, as we did before. Note that we get a warning about negative branch lengths in the NJ tree, which aren't allowed in the ML tree.

```{r ml} 
myseqs.phydat <- as.phyDat(myseqs)
myseqs.gtrig <- pml(mytree,myseqs.phydat,model="GTR+I+G",k=4)
myseqs.gtrig <- optim.pml(myseqs.gtrig,optNni=TRUE,optBf=TRUE,optQ=TRUE,optInv=TRUE,optGamma=TRUE,optEdge=TRUE,optRate=FALSE)
myseqs.mltree <- myseqs.gtrig$tree
```

We need to root the tree in order to do ancestral reconstruction. We could use ```rtt``` or ```lsd```, but in principle, we could use any method we discussed before. We scan the names of the tip labels, to get the tip dates and location.

```{r parse_tiplabels}
info <- scan(what=list(character(),character(),character(),character(),integer()),sep="_",quote="\"",text=paste(myseqs.mltree$tip.label,collapse="\n"),quiet=TRUE)
tipnames <- myseqs.mltree$tip.label
tipdates <- as.double(info[[5]])
tipdates
```

We can now root the tree and turn branch lengths into time using LSD.

```{r}
write.tree(myseqs.mltree,"H5N1.tre")
write.table(rbind(c(length(tipnames),""),cbind(tipnames,tipdates)),"H5N1.td",col.names=FALSE,row.names=FALSE,quote=FALSE)

```

```{r}
lsd.cmd <- sprintf("lsd -i %s -d %s -c -n 1 -r -b %s -s %s -v","H5N1.tre","H5N1.td",paste(10),seq.len)
lsd.cmd
lsd <- system(lsd.cmd,intern=TRUE)
```

```{r}
procresult <- function(fn){
  result <- readLines(fn)
  line <- result[grep("Tree 1 rate ",result)]
  line.split <- strsplit(line, " ")[[1]]
  list(rate=as.double(line.split[4]),tmrca=as.double(line.split[6]))
}
procresult("H5N1_result.txt")
```

```{r}
lsd.tree <- read.tree("H5N1_result_newick_date.txt")
```

Now we can extract the location, and reconstruct the changes in state. Here is another R trick to parse sequence headers.

```{r}
info <- scan(what=list(character(),character(),character(),character(),integer()),sep="_",quote="\"",text=paste(lsd.tree$tip.label,collapse="\n"),quiet=TRUE)
info
```

The location is the third entry in the list.

```{r}
mylocation <- as.factor(info[[3]])
mylocation
```

We fix any small branch lengths.

```{r }
lsd.tree$edge.length[lsd.tree$edge.length<0.00000001] <- 0.00000001
```

Now perform ancestral reconstruction of the location; I use a simple equal rates model (```model="ER"```), although in principle I could use more complex ones.

```{r}
lsd.tree.ace <- ace(mylocation,lsd.tree,type="discrete",method="ML",model="ER")
```

```{r } 
lsd.tree.ace
```

```{r } 
plot(lsd.tree, type="p",label.offset=0.0025,cex=0.75)
co <- c("blue", "yellow","red","green","orange")
tiplabels(pch = 22, bg = co[as.numeric(mylocation)], cex = 1.0)
nodelabels(thermo = lsd.tree.ace$lik.anc, piecol = co, cex = 0.25)
```

The next steps in this analysis could be:
- To explore different possible models for the migration process
- To perform a joint trait/tree analysis in BEAST (see [here](https://github.com/sdwfrost/influenza-dynamics-practical-h5n1) for an analysis of the same data)
- To develop more complex structured coalescent models e.g. in ```rcolgem```
