# 'Mugration' exercise

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Load the libraries.

```{r load_libraries} 
library(ape)
library(phangorn)
library(ggtree)
```

Load the FASTA data.

```{r load_data} 
myseqs <- read.dna("H5N1.fas",format="fasta")
```

```{r nj}
mytree  <- nj(dist.dna(myseqs,model="TN93"))
```

Starting with the neighbour joining tree, we reconstruct a maximum likelihood tree, as we did before. Note that we get a warning about negative branch lengths in the NJ tree, which aren't allowed in the ML tree.

```{r ml} 
myseqs.phydat <- as.phyDat(myseqs)
myseqs.gtrig <- pml(mytree,myseqs.phydat,model="GTR+I+G",k=4)
myseqs.gtrig <- optim.pml(myseqs.gtrig,optNni=TRUE,optBf=TRUE,optQ=TRUE,optInv=TRUE,optGamma=TRUE,optEdge=TRUE,optRate=FALSE)
myseqs.mltree <- myseqs.gtrig$tree
```

We need to root the tree in order to do ancestral reconstruction. We use ```rtt```, but in principle, we could use any method we discussed before. We scan the names of the tip labels, to get the tip dates and location.

```{r parse_tiplabels}
info <- scan(what=list(character(),character(),character(),character(),integer()),sep="_",quote="\"",text=paste(myseqs.mltree$tip.label,collapse="\n"),quiet=TRUE)
tipdates <- as.double(info[[5]])
tipdates
```

Now we can root with ```rtt```.

```{r}
myseqs.mltree.rooted <- rtt(myseqs.mltree,tipdates)
```

Now we can extract the location, and reconstruct the changes in state.

```{r}
info <- scan(what=list(character(),character(),character(),character(),integer()),sep="_",quote="\"",text=paste(myseqs.mltree.rooted$tip.label,collapse="\n"),quiet=TRUE)
mylocation <- as.factor(info[[3]])
mylocation
```


```{r }
myseqs.mltree.rooted$edge.length[myseqs.mltree.rooted$edge.length<0.00000001] <- 0.00000001
myseqs.ace <- ace(mylocation,myseqs.mltree.rooted,type="discrete",method="ML",model="ER")
```

```{r } 
myseqs.ace
```

```{r } 
plot(myseqs.mltree.rooted, type="p",label.offset=0.0025,cex=0.75)
co <- c("blue", "yellow","red","green","orange")
tiplabels(pch = 22, bg = co[as.numeric(mylocation)], cex = 1.0)
nodelabels(thermo = myseqs.ace$lik.anc, piecol = co, cex = 0.25)
```
