# Alignment example

## Prerequisites

First, load the libraries you need. We will use ```Biostrings```, which will allow us to translate sequences, ```msa```, which allows us to align using several external packages (bundled into R for easy installation), and ```seqinr``` for more sequence manipulation.

```{r}
library(Biostrings)
library(msa)
library(seqinr)
library(magrittr)
source("utils.R")
```

It's also useful to set any run-specific filenames at the beginning, so you can reuse the document more easily. For this example, I'll use the Ray et al. (2000) dataset, edited so that it isn't aligned.

```{r}
nucseq.fn <- "ray2000_edited.fas"
myref.fn <- "ray2000_ref.fas"
```

I'll use additional file stubs to keep track of what we've done to the original data.

## Align in nucleotide space

First, we will align in nucleotide space. This is advisable for non-coding data (e.g. the LTR of many viruses), and can be sufficient for many coding regions.

```{r}
nucseq <- readDNAStringSet(nucseq.fn)
nucseq
```

```{r}
nucseq.align <- msa(nucseq,method="Muscle")
```

Save the resulting alignment to a file.

```{r}
writeXStringSet(nucseq.align@unmasked,paste(nucseq.fn,".align",sep=""))
```
